---
title: "EDS 230 Homework 6: Using Sobol with an ODE"
format: pdf
author: Leela Dixit, Jay Kim, Richard Montes Lemus
execute:
  warning: false
  message: false
  freeze: auto
---

```{r setup, include=FALSE}
#| message: false
#| warning: false

# Load Libraries
library(tidyverse)
library(deSolve)
library(here)
library(sensitivity)
```

## Introduction

Forests accumulate carbon over time, but their growth dynamics can change once the canopy closes. In this assignment, we simulate forest carbon growth using an ODE model with two growth regimes: (1) exponential growth before canopy closure and (2) slower, density-limited growth after canopy closure. We first run the model with the provided parameter values and visualize the trajectory over 300 years. Next, we quantify the effect of uncertainty in key parameters on the maximum forest size using a global Sobol sensitivity analysis.

## Forest Growth Model Implementation

### Load Model Function

We source the differential equation function that defines forest growth (the derivative $\frac{dC}{dt}$) from an external R script. This allows us to use the ODE solver with our custom model.

```{r source-model}
#| message: false
#| warning: false

# Source population growth model
source(here("R/dforestgrowth.R"))

# Show the equation
print(dforestgrowth)
```

### Running the Model with Initial Parameters

Here, we define the initial forest carbon `Cinitial`, the simulation time vector (1-300 years), and the baseline parameter values provided. After initializing all the parameters, we will run the ODE solvers. The output is a time series of forest size.

```{r run-model}
#| message: false
#| warning: false

# Define starting forest size, simulation time, and model parameters
Cinitial <- 10 # Assume initial forest size is 10 kg/C
tm <- seq(from = 1, to = 300)
parms <- list(K = 250, r = 0.01, g = 2, thresh = 50)

# Run the ODE model for 300 years
ode <- ode(Cinitial, tm, dforestgrowth, parms)

# Rename columns for readability
colnames(ode) <- c("time", "forest_size")
head(ode)
tail(ode)
```

## Visualizing Forest Growth Trajectory

This plot shows the forest carbon trajectory over time under baseline parameters, including changes in behavior near the canopy closure threshold.

```{r growth-viz}
#| message: false
#| warning: false

# Visualize forest growth over time with fixed parameters
ggplot(as.data.frame(ode), aes(x = time, y = forest_size)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Time series of forest growth with canopy\nclosure threshold at 50 kgC",
       x = "Years",
       y = "Forest Size (Carbon)")
```

## Run a Sobol Global Sensitivity Analysis with Parameter Sampling

### Define Parameter Uncertainty & Generate Samples

We are interested in estimating the **maximum forest size** varies with parameter variation over a time period of 300 years. We will assume all parameters are normally distributed and generate two independent sample matrices (X1 and X2) for `sobolSalt()`. 

```{r sobol-sens}
#| message: false
#| warning: false

# Set the number of parameter samples you have as 2000
np <- 2000

# Generate parameters assuming normal distribution
K <- rnorm(mean = 250, sd = 25, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd = 0.2, n = np)
thresh <- rnorm(mean = 50, sd = 5, n = np)

# Combine sampled parameters into a data frame
X1 <- cbind.data.frame(r = r, K = K, g = g, thresh = thresh)

# Repeat to get our second set of samples
K <- rnorm(mean = 250, sd = 25, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd = 0.2, n = np)
thresh <- rnorm(mean = 50, sd = 5, n = np)

# Generate a second parameter sample set
X2 <- cbind.data.frame(r = r, K = K, g = g, thresh = thresh)

# Fix any negative values and they are not meaningful (using pmax and map_df)
X1 <- X1 %>% 
  map_df(pmax, 0.0)
X2 <- X2 %>% 
  map_df(pmax, 0.0)

# Create our Sobol object and get sets of
# Parameters for running the model, use `sobolSalt()`
sens_C <- sobolSalt(model = NULL, X1, X2, nboot = 300)

# Take a look at parameter sets
str(sens_C)

# Rename columns for readability
colnames(sens_C$X) <- c("r", "K", "g", "thresh")
head(sens_C$X)
```

## Computing the Model Metric

### Build a Wrapper Function to Run the Model and Compute a Metric

Our primary outcome metric is the maximum forest carbon achieved during the 300-year simulation and the wrapper takes one parameter set, solves the ODE, converts the output to a data frame, and returns the metric value. This is what we will map across all Sobol parameter sets.

```{r metric-wrapper}
#| message: false
#| warning: false

# Define simulation period for sensitivity analysis runs
simtimes <- seq(from = 1, to = 300)

# Use the first parameter set (or you could just pick a test value), 
# Not the list because ode wants parameters as a list, so name each
parmtest <- list(r = sens_C$X[1, "r"], 
                K = sens_C$X[1, "K"],
                g = sens_C$X[1, "g"],
                thresh = sens_C$X[1, "thresh"])
# Run ode
result <-  ode(y = Cinitial, 
               times = simtimes, 
               func = dforestgrowth, 
               parms = parmtest)

# Turn ode results into a labelled data frame 
result <- data.frame(time = result[ , 1], C = result[ , 2])

# Create function to pull out metrics we want 
compute_metrics <- function(result) {
  
  maxsize <- max(result$C)
  return(maxsize)
  
}

# Run metrics on results, lets use a threshold of 100
compute_metrics(result)

# Wrap ode solver and computing metrics 
p_wrapper <- function(r, K, g, 
                      thresh, 
                      Cinitial, 
                      simtimes, 
                      odefunc, 
                      metricfunc) {
  
  # Create parameter list
  parms <- list(r = r, K = K, g = g, thresh = thresh)
  
  # Run ode
  result <- ode(y = Cinitial, 
               times = simtimes, 
               func = odefunc, 
               parms = parms)
  
  # Results as a data frame
  result <- data.frame(time = result[ , 1], C = result[ , 2])
  
  # Get and compute metrics
  metric <- metricfunc(result)
  return(metric)
  
}

# Test with specific r and K values (r = 0.01, K = 150) and Cinitial = 10 
# Just do 10 time steps
p_wrapper(r = 0.01, K = 150, g = 2, thresh = 50,
          Cinitial = 10, 
          simtimes = seq(from = 1, to = 10),
          odefunc = dforestgrowth, 
          metricfunc = compute_metrics)

# Run model across all Sobol parameter combinations
allresults <- as.data.frame(sens_C$X) %>% 
  pmap(p_wrapper, 
       Cinitial = Cinitial, 
       simtimes = simtimes, 
       odefunc = dforestgrowth, 
       metricfunc = compute_metrics)

# Pull out data from results list
allres <- as.data.frame(allresults) %>% 
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "maxsize")
```

## Plot Sobol Sensitivity Analysis Results

### Evaluate the metric across all Sobol parameter sets

We run the wrapper over all Sobol parameter combinations to produce a distribution of maximum forest sizes.


```{r sobol-viz}
#| message: false
#| warning: false

# Create a box plot showing uncertainty in maximum forest size
# Across all Sobol parameter simulation
ggplot(allres, aes(x = "", y = maxsize)) + # Keep only one category
  geom_boxplot() + 
  coord_flip() +
  theme_minimal() +
  labs(title = "Uncertainty in Maximum Forest Size (300-year Maximum)",
       x = NULL,
       y = "Maximum Forest Size (kg C)") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Record Sobol Indices

We pass the computed metric values into `tell()` to estimate first-order (`S`) and total-order (`T`) Sobol indices.

```{r sobol-indices}
#| message: false
#| warning: false

# Compute Sobol sensitivity indices
sens_C_maxsize <- sensitivity::tell(sens_C, allres$maxsize)

# Label parameters in first-order Sobol Index
rownames(sens_C_maxsize$S) <- c("r", "K", "g", "thresh")
sens_C_maxsize$S

# Label parameters in total-order Sobol Index
rownames(sens_C_maxsize$T) <- c("r", "K", "g", "thresh")
sens_C_maxsize$T
```

## Conclusion:

In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about how what parameters climate change might influence).

Our sensitivity analysis shows that maximum forest size is most influenced by the carrying capacity (K) and early growth rate (r), these explain a majority of the variation in outcomes. Post-canopy growth rate (g) contributed moderately, while canopy closure threshold had little influence on how large the forest ultimately gets. This has large implications for climate change, as both K and r could be reduced by increasing extreme weather events - leading to a smaller forests.

## Extra Credit: Forest Size After 100 Years

### Define the second metric

We are going to compute Sobol indices for forest size at 100 years. Previously we simulated at integer years, we can extract `C` when time equals 100 and re-run the wrapper with year-100 metric.

```{r ec-compute}
#| message: false
#| warning: false

# Function to extract forest size after 100 years
compute_C100 <- function(result) {
  
  C100 <- result$C[result$time == 100]
  return(C100)
  
}

# Run sensitivity analysis using forest size at year 100
allresults_C100 <- as.data.frame(sens_C$X) %>% 
  pmap(p_wrapper, 
       Cinitial = Cinitial, 
       simtimes = simtimes, 
       odefunc = dforestgrowth, 
       metricfunc = compute_C100)

# Convert results into a tidy data frame for analysis
allres_C100 <- as.data.frame(allresults_C100) %>% 
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "C100")
```

### Visualizing Uncertainty in Forest Size at Year 100

This plot shows how forest size after 100 years varies across all Sobol parameter simulations. The box plot summarizes the distribution of outcomes to visualize the spread and variability caused by parameter uncertainty.

```{r ec-sobol-viz}
#| message: false
#| warning: false

# Visualize uncertainty in forest size at year 100
ggplot(allres_C100, aes(x = "", y = C100)) + # Keep only one category
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Uncertainty in Forest Size at 100 Years",
       x = NULL,
       y = "Maximum Forest Size (kg C)") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Computing Sobol Sensitivity Indices for Forest Size at Year 100

Here, we calculate Sobol sensitivity indices using forest size at 100 years as the response variable. The first-order indices (`S`) measure the individual contribution of each parameter to output variability, while the total-order indices (`T`) include both individual effects and parameter interactions.

```{r ec-sobol-indices}
#| message: false
#| warning: false

# Compute Sobol indices for the 100 year
sens_C_100 <- sensitivity::tell(sens_C, allres_C100$C100)

# Label parameters in first-order Sobol Index
rownames(sens_C_100$S) <- c("r", "K", "g", "thresh")
sens_C_100$S

# Label parameters in total-order Sobol Index
rownames(sens_C_100$T) <- c("r", "K", "g", "thresh")
sens_C_100$T
```

### Conclusion

In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about how what parameters climate change might influence).

Based on our Sobol indices for the max forest size at 100 years, we can see that almost all of the variation is due to the growth rate before canopy closure threshold. We can assume that external factors such as climate change that could influence max forest size after 100 years of growth will be almost entirely due to its influence on pre-canopy closure threshold growth rate.

