---
title: "HW7_twovariable_dynamic"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(here)
library(deSolve)
library(sensitivity)
library(ggpubr)
```

```{r}
source(here("R", "lotvmodK.R"))

print(lotvmodK)
```

```{r}
# initial conditions
currpop <- c(prey = 10, pred = 1)

# set parameter list
pars <- c(rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.4, K = 2000, hunt = 0.5, threshold = 20)

# times when you want to evaluate
days <- seq(from = 1, to = 500)

# run our differential equation solver
res <- ode(func = lotvmodK, y = currpop, times = days, parms = pars)

# rearrange for plotting
resl <- as.data.frame(res) %>% pivot_longer(-time, names_to = "species", values_to = "pop")

# graph both populations over time
p1 <- ggplot(resl, aes(time, pop, col = species)) +
  geom_line()
p1

# also look at relationships between preditor and prey population and use color for time
# I will remove the legend here to make it easier to see
p2 <- ggplot(as.data.frame(res), aes(pred, prey, col = (round(time / 10)))) +
  geom_point() +
  theme(legend.position = "none")
p2
p2 <- ggplot(as.data.frame(res), aes(pred, prey, col = as.factor(round(time / 10)))) +
  geom_point() +
  theme(legend.position = "none")
p2
ggarrange(p1, p2)
```

Sobol Sensitivity Testing

```{r}
# want to learn about sensitivity to growth rate (r) and carrying capacity
# set the number of parameters
np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)
hunt <- runif(min = 0.01, max = 1, n = np)
threshold <- runif(min = 1, max = 300, n = np)

X1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, hunt = hunt, thresh = threshold)

# repeat to get our second set of samples
np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)
hunt <- runif(min = 0.01, max = 1, n = np)
threshold <- runif(min = 1, max = 300, n = np)

X2 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, hunt = hunt, thresh = threshold)

# create our sobel object and get sets of parameters for running the model
sens_PP <- sobolSalt(model = NULL, X1, X2, nboot = 300)

# name parameter sets...
colnames(sens_PP$X) <- c("rprey", "K", "alpha", "eff", "pmort", "hunt", "threshold")

# our metrics
# lets say we  want the maximum and minimum  of both predator and prey

compute_metrics <- function(result) {
  maxprey <- max(result$prey)
  minprey <- min(result$prey)
  return(list(maxprey = maxprey, minprey = minprey))
}



# build a wrapper function

p_wrapper <- function(rprey, alpha, eff, pmort, K, hunt, currpop, threshold, days, func) {
  parms <- list(rprey = rprey, alpha = alpha, eff = eff, pmort = pmort, K = K, hunt = hunt, threshold = threshold)
  result <- ode(y = currpop, times = days, func = func, parms = parms)
  colnames(result) <- c("time", "prey", "pred")
  # get metrics
  metrics <- compute_metrics(as.data.frame(result))
}
```

```{r}
# run our model for all parameters and extract the results
currpop <- c(prey = 1, pred = 1)
days <- seq(from = 1, to = 500)
allresults <- as.data.frame(sens_PP$X) %>% 
    pmap(p_wrapper, currpop = currpop, days = days, func = lotvmodK)

np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)
hunt <- runif(min = 0.01, max = 1, n = np)
threshold <- runif(min = 1, max = 50, n = np)

# take results back to unlisted form
allres <- allresults %>% map_dfr(`[`, c("maxprey", "minprey"))

# range of response across parameter uncertainty
allresl <- allres %>% gather(key = "metric", value = "pop")
ggplot(allresl, aes(metric, pop)) +
  geom_boxplot()

# dealing with different scales
ggplot(allresl, aes(metric, pop, col = metric)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free")
# plot cummulative densities

ggplot(allresl, aes(pop, col = metric))  +
  stat_ecdf(geom = "line") +
  facet_wrap(~metric, scales = "free")
```

# Part 2

## Exploring how different hunting levels and uncertainty/variation in parameters in our model affect the population stability

1) Use Sobol to sample a variety parameters and hunting rates

```{r}
# Create first set of parameter and hunting samples 
np <- 200 # Number of parameter sets 

# Sample biological parameters based on best guess provided in instructions
# Use best guess as the mean with small SD to represent parameter uncertainty
K <- rnorm(mean = 2000, sd = 200,  n = np)  
rprey <- rnorm(mean = 0.95, sd = 0.1,  n = np)  
alpha <- rnorm(mean = 0.01, sd = 0.001,n = np)  
eff <- rnorm(mean = 0.6,  sd = 0.06, n = np)  
pmort <- rnorm(mean = 0.4,  sd = 0.04, n = np)  


# Sample hunt uniformly from 0 to 0.5 — no best guess so full range explored
# Capped at 0.5 because removing more than 50% of prey per day is unrealistic
hunt <- runif(min = 0.0, max = 0.5,   n = np) 

# Keep threshold fixed to isolate affect of hunt
threshold <- 20

# Create first table with parameter and hunting samples
prey_pop_parameters1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, hunt = hunt, thresh = rep(20, np))


# Create second set of parameter and hunting samples 
np <- 200 # Number of parameter sets 

# Sample biological parameters based on best guess provided in instructions
# Use best guess as the mean with small SD to represent parameter uncertainty
K <- rnorm(mean = 2000, sd = 200,  n = np)  
rprey <- rnorm(mean = 0.95, sd = 0.1,  n = np)  
alpha <- rnorm(mean = 0.01, sd = 0.001,n = np)  
eff <- rnorm(mean = 0.6,  sd = 0.06, n = np)  
pmort <- rnorm(mean = 0.4,  sd = 0.04, n = np)  


# Sample hunt uniformly from 0 to 0.5, no best guess so full range explored
# Capped at 0.5 because removing more than 50% of prey per day is unrealistic
hunt <- runif(min = 0.0, max = 0.5,   n = np) 

# Keep threshold fixed to isolate affect of hunt
threshold <- 20

# Create second table with parameter and hunting samples
prey_pop_parameters2 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, hunt = hunt, thresh = rep(20, np))

# Use two tables to combine different parameter and hunting rates that will 
# be used for running the model   
sens_PP <- sobolSalt(model = NULL, prey_pop_parameters1, prey_pop_parameters2, nboot = 100)   

# Name parameter sets
colnames(sens_PP$X) <- c("rprey", "K", "alpha", "eff", "pmort", "hunt", "threshold")

```


2) Compute metric for stability and create wrapper function 
```{r}

# Create a metric function to define stability 
compute_metrics <- function(result) {
  result <- as.data.frame(result)
  final_prey <- tail(result$prey, 1)
  
# Stability is defined as having an ending prey population above 50
   if (final_prey > 50) { 
     return(list(final_prey = final_prey))
   } else {
     return(list(final_prey = NA)) # Return NA if prey population is below 50
   }
}

# Wrapper function to run the ODE model for a single parameter set and return stability metric (assigned as stable or unstable)
p_wrapper <- function(rprey, alpha, eff, pmort, K, hunt, currpop, threshold, days, func) {
  parms <- list(rprey = rprey, alpha = alpha, eff = eff, pmort = pmort, K = K, 
                hunt = hunt, threshold = threshold)
  result <- ode(y = currpop, times = days, func = func, parms = parms)
  colnames(result) <- c("time", "prey", "pred")
  metrics <- compute_metrics(as.data.frame(result))
  return(metrics)
}

```

3) Use parameter sets to predict prey population and assess stability

```{r, warning=FALSE, message=FALSE, results='hide'}
# Run our model for all parameters and extract the results
currpop <- c(prey = 200, pred = 5) # Define starting prey and pred population
days <- seq(from = 1, to = 500)
allresults <- as.data.frame(sens_PP$X) %>% 
    pmap(p_wrapper, currpop = currpop, days = days, func = lotvmodK)

```

```{r}
# Put model results in data frame 
allres <- allresults %>% map_dfr(`[`, c("final_prey"))

# Bind parameter sets to results
final_table <- bind_cols(as.data.frame(sens_PP$X), allres)

# Filter to only rows where parameter set values led to stability
stable_only <- final_table %>% filter(!is.na(final_prey))

stable_only
```

4) Compare parameter distributions between stable and unstable runs

```{r}
# Create separate box plot per parameter
final_table %>%
  mutate(stable = !is.na(final_prey)) %>%
  pivot_longer(cols = c(rprey, K, alpha, eff, pmort, hunt),
               names_to = "parameter", values_to = "value") %>%
  ggplot(aes(stable, value, fill = stable)) +
  geom_boxplot() +
  facet_wrap(~parameter, scales = "free_y") +
  labs(title = "Parameter distributions: stable vs unstable runs")
```

### Discussion: 

**b.your exploration (e.g how you tested different hunting levels and how you defined a stability metric**

  I tested different hunting levels by including in my sobol a uniform sampling of values for hunting rate. This ranged from 0 - 0.5. I made sure to cap the hunting rate at 0.5 because removing more than 50% of prey per day is unrealistic. Each of these hunting levels is associated with a stable or unstable population. 
  
  I defined a stability metric by modifying my compute metric function so it changed the value in final_prey column to NA if the prey population at the end of the simulation was less than 50. If the final_prey value was above 50 it was left alone. I chose this stability metric because setting 50 as the prey population threshold for stability seems large enough to sustain prey reproduction. I did not use predator population in this stability metric because I thought prey population would capture predator population well since predators rely on prey as a food source. 
  
**c. How you estimated sustainable hunting level and a brief explanation of why you chose the metric you did**  

I estimated a sustainable hunting level by plotting the distribution of hunting rate for all parameter sets labeled as stable vs unstable. For parameter sets that led to a stable population, meaning the prey population was above 50 at the end of the simulation, the hunting rate was a bit below 0.2. Therefore, I'd recommend keeping hunting at or below 0.2 to maintain a viable prey population.

I chose boxplots comparing stable vs unstable runs because they let me see at a glance how each parameter differs between the two groups. This was useful for confirming that hunt was the main driver of stability and not some other parameter like K or alpha, which showed much less separation between stable and unstable runs.


