---
title: "EDS 230 Homework 7: Two Variables Dynamic"
format: pdf
author: Leela Dixit, Jay Kim, Richard Montes Lemus
execute:
  warning: false
  message: false
  freeze: auto
---

```{r setup, include=FALSE}
#| message: false
#| warning: false
#| code-fold: true

library(tidyverse)
library(here)
library(deSolve)
library(sensitivity)
library(ggpubr)
```

# Part 1: Building the Hunting Model

## Loading Modified Predator-Prey Model

This model extends the classic predator-prey equations by adding `hunt` (the hunting rate applied to prey) and `threshold` (the minimum prey population required before hunting is permitted). Hunting only occurs when prey exceeds the threshold, ensuring we never hunt more prey than exist.

```{r load-model}
#| message: false
#| warning: false

source(here("R", "lotvmodK.R"))
print(lotvmodK)
```

## Running the Baseline Model

Here we run the model with the best-guess parameters to verify behavior before the sensitivity analysis: `rprey = 0.95`, `alpha = 0.01`, `eff = 0.6`, `pmort = 0.4`, `K = 2000`, with a moderate hunting rate of 0.5 and a threshold of 20.

```{r baseline-setup}
#| message: false
#| warning: false

# Initial conditions
currpop <- c(prey = 10, pred = 1)

# Set parameter list
pars <- c(rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.4, K = 2000, hunt = 0.5, threshold = 20)

# Times when you want to evaluate
days <- seq(from = 1, to = 500)

# Run our differential equation solver
res <- ode(func = lotvmodK, y = currpop, times = days, parms = pars)

# Rearrange for plotting
resl <- as.data.frame(res) %>% pivot_longer(-time, names_to = "species", values_to = "pop")
```

## Visualizing Baseline Population Dynamics

The time series shows how both populations evolve over 500 days. The phase-plane plot reveals whether the system settles into a stable cycle or not. 

```{r baseline-plots}
#| message: false
#| warning: false

# Graph both populations over time
p1 <- ggplot(resl, aes(time, pop, col = species)) +
  geom_line()
p1

# Also look at relationships between preditor and prey population and use color for time
# I will remove the legend here to make it easier to see
p2 <- ggplot(as.data.frame(res), aes(pred, prey, 
                                     col = (round(time / 10)))) +
  geom_point() +
  theme(legend.position = "none")
p2

p2 <- ggplot(as.data.frame(res), aes(pred, prey, 
                                     col = as.factor(round(time / 10)))) +
  geom_point() +
  theme(legend.position = "none")
p2

ggarrange(p1, p2)
```

## Preliminary Sobol Sensitivity Analysis (Exploratory)

Before the focused Part 2 analysis, we run a broad Sobol sensitivity across all parameters to understand which drive prey population outcomes most. Parameters are sampled over wide exploratory ranges.

### Setting Up Parameter Samples

```{r sobol-setup-part-1}
#| message: false
#| warning: false

# Want to learn about sensitivity to growth rate (r) and carrying capacity
# Set the number of parameters
np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)
hunt <- runif(min = 0.01, max = 1, n = np)
threshold <- runif(min = 1, max = 300, n = np)

X1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, hunt = hunt, thresh = threshold)

# Repeat to get our second set of samples
np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)
hunt <- runif(min = 0.01, max = 1, n = np)
threshold <- runif(min = 1, max = 300, n = np)

X2 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, hunt = hunt, thresh = threshold)

# Create our sobel object and get sets of parameters for running the model
sens_PP <- sobolSalt(model = NULL, X1, X2, nboot = 300)

# Name parameter sets...
colnames(sens_PP$X) <- c("rprey", "K", "alpha", "eff", 
                         "pmort", "hunt", "threshold")
```

### Metrics and Wrapper Functions

`compute_metrics` extracts max and min prey population from a model run. `p_wrapper` runs the ODE for a single parameter set and returns these metrics.

```{r sobol-metric-function}
#| message: false
#| warning: false

# For our metrics we want the maximum and minimum of both predator and prey
compute_metrics <- function(result) {
  
  maxprey <- max(result$prey)
  minprey <- min(result$prey)
  return(list(maxprey = maxprey, minprey = minprey))
  
}

# Build a wrapper function
p_wrapper <- function(rprey, alpha, eff, pmort, K, hunt, 
                      currpop, threshold, days, func) {
  
  parms <- list(rprey = rprey, alpha = alpha, eff = eff, 
                pmort = pmort, K = K, hunt = hunt, threshold = threshold)
  result <- ode(y = currpop, times = days, func = func, parms = parms)
  colnames(result) <- c("time", "prey", "pred")
  
  # Get metrics
  metrics <- compute_metrics(as.data.frame(result))
  
}
```

### Results

```{r sobol-run-part-1}
#| message: false
#| warning: false
#| eval: true
#| echo: true
#| output: false

# Run our model for all parameters and extract the results
currpop <- c(prey = 1, pred = 1)
days <- seq(from = 1, to = 500)
allresults <- as.data.frame(sens_PP$X) %>% 
    pmap(p_wrapper, currpop = currpop, days = days, func = lotvmodK)

np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)
hunt <- runif(min = 0.01, max = 1, n = np)
threshold <- runif(min = 1, max = 50, n = np)

# Take results back to unlisted form
allres <- allresults %>% map_dfr(`[`, c("maxprey", "minprey"))
```

```{r result-plots}
#| message: false
#| warning: false

# Range of response across parameter uncertainty
allresl <- allres %>% gather(key = "metric", value = "pop")
ggplot(allresl, aes(metric, pop)) +
  geom_boxplot()

# Dealing with different scales
ggplot(allresl, aes(metric, pop, col = metric)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free")

# Plot cumulative densities
ggplot(allresl, aes(pop, col = metric))  +
  stat_ecdf(geom = "line") +
  facet_wrap(~metric, scales = "free")
```


# Part 2: Exploring Hunting Levels and Population Stability

Here we formally explore how hunting rate, combined with biological parameter uncertainty, affects long-term stability. We use Sobol sampling to test many combinations and apply a stability metric to each run. We will explore how different hunting levels and uncertainty/variation in parameters in our model affect the population stability.

## 1) Use Sobol to Sample a Variety Parameters and Hunting Rates

Biological parameters are sampled with normal distributions centered on the best-guess values (SD = 10% of mean). Hunting rate is sampled uniformly from 0–0.5; capped at 0.5 since removing more than 50% of prey per day is ecologically unrealistic. Threshold is fixed at 20 to isolate the hunting rate effect.

```{r sobol-setup-part-2}
#| message: false
#| warning: false

# Create first set of parameter and hunting samples 
np <- 200 # Number of parameter sets 

# Sample biological parameters based on best guess provided in instructions
# Use best guess as the mean with small SD to represent parameter uncertainty
K <- rnorm(mean = 2000, sd = 200, n = np)  
rprey <- rnorm(mean = 0.95, sd = 0.1, n = np)  
alpha <- rnorm(mean = 0.01, sd = 0.001, n = np)  
eff <- rnorm(mean = 0.6, sd = 0.06, n = np)  
pmort <- rnorm(mean = 0.4, sd = 0.04, n = np)  

# Sample hunt uniformly from 0 to 0.5 — no best guess so full range explored
# Capped at 0.5 because removing more than 50% of prey per day is unrealistic
hunt <- runif(min = 0.0, max = 0.5, n = np) 

# Keep threshold fixed to isolate affect of hunt
threshold <- 20

# Create first table with parameter and hunting samples
prey_pop_parameters1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, hunt = hunt, thresh = rep(20, np))


# Create second set of parameter and hunting samples 
np <- 200 # Number of parameter sets 

# Sample biological parameters based on best guess provided in instructions
# Use best guess as the mean with small SD to represent parameter uncertainty
K <- rnorm(mean = 2000, sd = 200,  n = np)  
rprey <- rnorm(mean = 0.95, sd = 0.1,  n = np)  
alpha <- rnorm(mean = 0.01, sd = 0.001,n = np)  
eff <- rnorm(mean = 0.6,  sd = 0.06, n = np)  
pmort <- rnorm(mean = 0.4,  sd = 0.04, n = np)  

# Sample hunt uniformly from 0 to 0.5, no best guess so full range explored
# Capped at 0.5 because removing more than 50% of prey per day is unrealistic
hunt <- runif(min = 0.0, max = 0.5, n = np) 

# Keep threshold fixed to isolate affect of hunt
threshold <- 20

# Create second table with parameter and hunting samples
prey_pop_parameters2 <- cbind.data.frame(rprey = rprey, K = K, 
                                         alpha = alpha, eff = eff, 
                                         pmort = pmort, hunt = hunt, 
                                         thresh = rep(20, np))

# Use two tables to combine different parameter and hunting rates that will 
# Be used for running the model   
sens_PP <- sobolSalt(model = NULL, prey_pop_parameters1, 
                     prey_pop_parameters2, nboot = 100)   

# Name parameter sets
colnames(sens_PP$X) <- c("rprey", "K", "alpha", "eff", 
                         "pmort", "hunt", "threshold")

```

## 2) Stability Metric and Wrapper Function 

**Stability metric:** A run is stable if the final prey population at day 500 exceeds 50. Below that, the run returns `NA` (unstable). We focus on prey because predator persistence depends on prey survival — if prey persist above 50, the ecosystem is likely viable.

```{r stability-metric}
#| message: false
#| warning: false

# Create a metric function to define stability 
compute_metrics <- function(result) {
  
  result <- as.data.frame(result)
  final_prey <- tail(result$prey, 1)
  
# Stability is defined as having an ending prey population above 50
   if (final_prey > 50) { 
     
     return(list(final_prey = final_prey))
     
   } else {
     
     return(list(final_prey = NA)) # Return NA if prey population is below 50
     
   }
  
}

# Wrapper function to run the ODE model for a single parameter set & 
# Return stability metric (assigned as stable or unstable)
p_wrapper <- function(rprey, alpha, eff, pmort, K, hunt, 
                      currpop, threshold, days, func) {
  
  parms <- list(rprey = rprey, alpha = alpha, eff = eff, 
                pmort = pmort, K = K, hunt = hunt, threshold = threshold)
  result <- ode(y = currpop, times = days, func = func, parms = parms)
  colnames(result) <- c("time", "prey", "pred")
  metrics <- compute_metrics(as.data.frame(result))
  return(metrics)
  
}
```

## 3) Running the Model Across All Parameter Sets

```{r sobol-run-part-2, warning=FALSE, message=FALSE, results='hide'}
#| message: false
#| warning: false
#| eval: true
#| echo: true
#| output: false

# Run our model for all parameters and extract the results
currpop <- c(prey = 200, pred = 5) # Define starting prey and pred population
days <- seq(from = 1, to = 500)
allresults <- as.data.frame(sens_PP$X) %>% 
    pmap(p_wrapper, currpop = currpop, days = days, func = lotvmodK)
```

```{r sobol-results-table}
#| message: false
#| warning: false

# Put model results in data frame 
allres <- allresults %>% map_dfr(`[`, c("final_prey"))

# Bind parameter sets to results
final_table <- bind_cols(as.data.frame(sens_PP$X), allres)

# Filter to only rows where parameter set values led to stability
stable_only <- final_table %>% filter(!is.na(final_prey))
stable_only
```

## 4) Compare Parameter Distributions between Stable and Unstable Runs

These boxplots compare each parameter between stable and unstable runs. Parameters with minimal overlap between groups are the strongest drivers of stability.

```{r stability-boxplots}
#| message: false
#| warning: false

# Create separate box plot per parameter
final_table %>%
  mutate(stable = !is.na(final_prey)) %>%
  pivot_longer(cols = c(rprey, K, alpha, eff, pmort, hunt),
               names_to = "parameter", values_to = "value") %>%
  ggplot(aes(stable, value, fill = stable)) +
  geom_boxplot() +
  facet_wrap(~parameter, scales = "free_y") +
  labs(title = "Parameter distributions: stable vs unstable runs")
```

### Discussion: 

**b.your exploration (e.g how you tested different hunting levels and how you defined a stability metric**

  I tested different hunting levels by including in my sobol a uniform sampling of values for hunting rate. This ranged from 0 - 0.5. I made sure to cap the hunting rate at 0.5 because removing more than 50% of prey per day is unrealistic. Each of these hunting levels is associated with a stable or unstable population. 
  
  I defined a stability metric by modifying my compute metric function so it changed the value in final_prey column to NA if the prey population at the end of the simulation was less than 50. If the final_prey value was above 50 it was left alone. I chose this stability metric because setting 50 as the prey population threshold for stability seems large enough to sustain prey reproduction. I did not use predator population in this stability metric because I thought prey population would capture predator population well since predators rely on prey as a food source. 

**c. How you estimated sustainable hunting level and a brief explanation of why you chose the metric you did**  

I estimated a sustainable hunting level by plotting the distribution of hunting rate for all parameter sets labeled as stable vs unstable. For parameter sets that led to a stable population, meaning the prey population was above 50 at the end of the simulation, the hunting rate was a bit below 0.2. Therefore, I'd recommend keeping hunting at or below 0.2 to maintain a viable prey population.

I chose boxplots comparing stable vs unstable runs because they let me see at a glance how each parameter differs between the two groups. This was useful for confirming that hunt was the main driver of stability and not some other parameter like K or alpha, which showed much less separation between stable and unstable runs.